{RILS}-{ROLS} : Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-12-03 paper_resources/doc/bmc_jbd/bmc_jbd.tex
Final revisions almost done

\subsection{Datasets and \texttt{SRBench}}\fxnote{Marko: Dosao dovde!}

@@

Some \textsc{Random} instances are shown in Table~\ref{table:random}, while all instances can be found at \textsc{RILS-ROLS} GitHub repository: \url{https://github.com/kartelj/rils-rols}.

@@

\subsection{Comparison to other methods}
% ACA: ima 14 uporednih algoritama, ne 13
In this section, we evaluate our \textsc{RILS-ROLS} algorithm and compare it to 14 other competitors from the literature. The numerical results are reported in Tables~\ref{tab:comp_noise0}-\ref{tab:comp_noise001}. The results for \textsc{Feynman} and \textsc{Storgatz} are given w.r.t.\  the following levels of noise: 0.0 (no-noisy data), 0.001 (low-level noise), and 0.01 (high-level noise). More precisely, the white Gaussian noise is added to the target variable the same way as in~\cite{la2021contemporary}:
$$ y_{noise} = y + \epsilon, \epsilon \sim \mathcal{N}\left(0, \alpha \sqrt{\frac{1}{n} \sum _{i=1}^n{y_i^2}}\right),$$
where $n$ relates to the number of items in the input data.

@@

\begin{itemize}
	\item Concerning the exact percentages in case when no noise is utilized in the input data, the best performing algorithm is our \textsc{RILS-ROLS}. It found the exact solution in 57.84\% of runs for \textsc{Feynman} instances. It was even more impressive in solving \textsc{Strogatz} instances, being successful in 83.57\% runs. The second best performing  approach is \textsc{AI-Feynman}, successful in 55.78\% runs for \textsc{Feynman} set, and 27.14\% runs for the problem instances from benchmark set \textsc{Strogatz}. All other approaches perform significantly worse, and none of them is able to deliver the exact solution for more than 30\% runs over the instances of any of the two considered ground-truth benchmark sets. The overall results are in favor of \textsc{RILS-ROLS} with 60.62\%, while the second best is \textsc{AI-Feynman} with 52.65\%. 
	\item  Concerning the exact results in presence of the small noise (0.001), our \textsc{RILS-ROLS} is still performing very well, finding an exact model in 42.08\% of runs considering all problem instances. The second best approach is again \textsc{AI-Feynman} being successful in 31.89\% runs. As in no-noise scenario, there is a significant difference in \textsc{AI-Feynman} performances among two sets of instances -- it performs much better on \textsc{Feynman} instances than on \textsc{Strogatz}. It is also worth mentioning that \textsc{AFP-FE} shows to be robust w.r.t. noise so its performances only slightly deteriorate in comparison to the no-noise scenario. This puts it on the third place with overall accuracy of 21.23\%. Other approaches reach accuracy bellow 20\%.  
	\item  Concerning the scenario with high noise (level of 0.01), our \textsc{RILS-ROLS} is still performing quite well -- it finds and exact model in 34.77\% runs. The second best is \textsc{AFP-FE} reaching 20\% correct solutions. \textsc{AI-Feynman} performances degrade significantly under this level of noise, so it is ranked only 6th overall, reaching the accuracy of 12.61\%.

@@

	\item When comparing performances of the algorithms in terms of $R^2 > 0.999$ accuracy, in no-noise scenario, our \textsc{RILS}-\textsc{ROLS} is successful with the rate of success of 83.38\% (third place). Better methods in terms of this accuracy are \textsc{MRGP}, and \textsc{Operon}, with 92.69\%, and 86.92\%, respectively. On the other hand, the exact model percentages of these two approaches are rather low (0\% and 16\%, respectively). These two algorithms are therefore more appropriate in scenarios when the underlying (ground-truth) model is unknown, so-called \emph{black-box} regression.

@@

\begin{itemize}
	\item When considering the instance problems with no-noise, the lowest average runtime is reported for \textsc{RILS}-\textsc{ROLS} algorithm (228 seconds). A slightly worse runtime is obtained by \textsc{DSR} algorithm (232 seconds). Note that the relative exact percentage produced by \textsc{RILS}-\textsc{ROLS} is far better from \textsc{DSR} (60.62\% versu 19.71\%). All remaining approaches deliver an order of magnitude larger average runtimes than that of \textsc{RILS}-\textsc{ROLS} and \textsc{DSR} methods..   
	\item  When considering instances with the noise level of 0.001, the lowest average runtime to find the exact model is again delivered by \textsc{RILS}-\textsc{ROLS} algorithm (108 seconds). The second best is produced by \textsc{DSR}, but this time significantly larger (622 seconds) compared to \textsc{RILS-ROLS}. 
	\item When considering the instance problems with the noise level of 0.01, the conclusion is similar to previous. 
\end{itemize}

@@

First, the omnibus rank-based Friedman’s test for all approaches was conducted. 
The number of ranks is 1300, i.e., for each run of each method, the separate rank is calculated. Rank is based on the value of indicator variable, which takes 1 if the method delivers correct model in a given run, otherwise it takes value of 0, no matter if method finished and delivered incorrect model, or method did not finish at all.   Further, the ranks are averaged and used to test the null hypothesis, which states that there is no statistical differences between the average ranks of all competitors. 
In the case in which the null hypothesis $H_0$ was rejected, pairwise comparisons were further performed by using the Nemenyi post-hoc test~\cite{pohlert2014pairwise}. The outcome is represented by means of critical difference (CD) plots. 
% ACA: ima 15 ukupno algoritama, ne 14
In each CD plot, the 15 competitors are placed on the horizontal axis according to their average ranking. Thereafter, the CD score is computed for a significance level of 0.05. If the difference is small enough, meaning that no statistical difference is detected, a horizontal bar linking statistically equal approaches is drawn. This analysis is done for all three levels of noise, and the corresponding CD plots are shown in Figures~\ref{fig:CDplots-no-noise}-\ref{fig:CDplots-noise0.01}.

@@

In this section we study the scalability of \textsc{RILS-ROLS} w.r.t. size of instances problems under different levels of noise.
Since we do not compare \textsc{RILS-ROLS} to other competitors here, but just assess the method scalability and sensitivity to noise, the experimental setup is different. First, we run each instance only once -- note that are 5 randomly generated instances for each of 47 combinations of (size, variable count). Second, the exit criteria are different: 1) maximal running time is 300 seconds instead of 1 hour, and 2) maximal number of fitness evaluations is 100 thousands instead of 1 million. 
As before, test set takes 25\% of total data. 
For the sake of having simpler presentation, we split the instances from \textsc{Random} benchmark into three parts as follows. 
\begin{itemize}
	\item \textit{Small-sized instances}: instances with solution size from 3 to 7.
	\item \textit{Medium-sized instances}: those with size from 8 to 11.
	\item \textit{Large-sized instances}: instances with size from 12 to 15. 
\end{itemize}

@@

	\item We can conclude that increased formula size and noise level jointly contribute to deterioration of overall \textsc{RILS-ROLS} accuracy, which is expected and natural. The diagram suggests that this deterioration behaves similar across different levels of noise when instance size changes from small to medium -- almost linearly, and with almost exact slope. On the other hand, the relative deterioration under no-noise level, when going from medium to large instance is higher than under noisy data. Possible explanation for this is that the exact solution accuracy of medium instances is already very deteriorated in noisy scenario, therefore, further increase in size does not have significant impact. 
\end{itemize}

@@

The following conclusions may be drawn from these results.

>>>>

\subsection{Datasets and \texttt{SRBench}}

@@

Some random instances are shown in Table~\ref{table:random}, while all instances can be found at %\textsc{RILS-ROLS} GitHub 
 the following repository: \url{https://github.com/kartelj/rils-rols}.

@@

\subsection{Comparison to other methods}
% ACA: ima 14 uporednih algoritama, ne 13
In this section, we evaluate our \textsc{RILS-ROLS} algorithm and compare it to 14 other competitors from the literature. The numerical results are reported in Tables~\ref{tab:comp_noise0}-\ref{tab:comp_noise001}. The results for \textsc{Feynman} and \textsc{Storgatz} benchmark are given w.r.t.\  the following levels of noise: 0.0 (no-noisy data), 0.001 (low-level noise), and 0.01 (high-level noise). More precisely, the white Gaussian noise is added to the target variable the same way as in~\cite{la2021contemporary}:
$$ y_{noise} = y + \epsilon, \epsilon \sim \mathcal{N}\left(0, \alpha \sqrt{\frac{1}{n} \sum _{i=1}^n{y_i^2}}\right),$$
where $n$ relates to the number of items in the input data.

@@

\begin{itemize}
	\item Concerning the exact percentages in case when no noise is utilized in the input data, the best performing algorithm is our \textsc{RILS-ROLS}. It found the exact solution in 57.84\% of runs for \textsc{Feynman} instances. It is even more impressive in solving \textsc{Strogatz} instances, being successful in 83.57\% runs. The second best performing  approach is \textsc{AI-Feynman}, successful in 55.78\% runs for \textsc{Feynman} set, and 27.14\% runs for the problem instances from benchmark set \textsc{Strogatz}. All other approaches perform significantly worse, and none of them is able to deliver the exact solution for more than 30\% runs over the instances of any of the two considered ground-truth benchmark sets. The overall results are in favor of \textsc{RILS-ROLS} with 60.62\%, while the second best is \textsc{AI-Feynman} with 52.65\%. 
	\item  Concerning the exact results in presence of the small noise (0.001), our \textsc{RILS-ROLS} is still performing   well, finding an exact model in 42.08\% of runs considering all problem instances. The second best approach is again \textsc{AI-Feynman} being successful in 31.89\% runs. As in no-noise scenario, there is a significant difference in \textsc{AI-Feynman} performances among two sets of instances -- it performs much better on \textsc{Feynman} instances than on \textsc{Strogatz}. It is also worth mentioning that \textsc{AFP-FE} shows to be robust w.r.t. noise so its performances only slightly deteriorate in comparison to the no-noise scenario. This puts it on the third place with overall accuracy of 21.23\%. Other approaches reach accuracy bellow 20\%.  
	\item  Concerning the scenario with high noise (level of 0.01), our \textsc{RILS-ROLS} is still performing quite well -- it finds and exact model in 34.77\% runs. The second best is \textsc{AFP-FE} reaching 20\% correct solutions. \textsc{AI-Feynman} performances degrade significantly under this level of noise, so it is ranked only 6th overall, reaching the accuracy of 12.61\%.

@@

	\item When comparing performances of the algorithms in terms of $R^2 > 0.999$ accuracy, in no-noise scenario, our \textsc{RILS}-\textsc{ROLS} is successful with the rate of success of 83.38\% (ranked on the third place). Better methods in terms of this accuracy are \textsc{MRGP}, and \textsc{Operon}, with 92.69\%, and 86.92\%, respectively. On the other hand, the exact model percentages of these two approaches are rather low (0\% and 16\%, respectively). These two algorithms are therefore more appropriate in scenarios when the underlying (ground-truth) model is unknown, so-called \emph{black-box} regression.

@@

\begin{itemize}
	\item When considering the instance problems with no-noise, the lowest average runtime is reported for \textsc{RILS}-\textsc{ROLS} algorithm (228 seconds). A slightly worse runtime is obtained by \textsc{DSR} algorithm (232 seconds). Note that the relative exact percentage produced by \textsc{RILS}-\textsc{ROLS} is far better from \textsc{DSR} (60.62\% versus 19.71\%). All remaining approaches deliver an order of magnitude larger average runtimes than that of \textsc{RILS}-\textsc{ROLS} and \textsc{DSR} methods..   
	\item  When considering instances with the noise level of 0.001, the lowest average runtime to obtain the exact model is again delivered by \textsc{RILS}-\textsc{ROLS} algorithm (108 seconds). The second best is produced by \textsc{DSR}, but this time significantly larger (622 seconds) compared to \textsc{RILS-ROLS}. 
	\item When considering the instance problems with the noise level of 0.01, the conclusion is similar to previous. 
\end{itemize}

@@

First, the omnibus rank-based Friedman’s test for all approaches was conducted. 
The number of ranks is 1300, i.e., for each run of each method, the separate rank is calculated. Rank is based on the value of indicator variable, which takes 1 if the method delivers correct model in a given run, otherwise it takes value of 0, no matter if method finished and delivered incorrect model, or the method did not finish at all.   Further, the ranks are averaged and used to test the null hypothesis, which states that there is   statistical differences between the average ranks of all competitors. 
In the case the null hypothesis $H_0$ was rejected, pairwise comparisons were further performed by using the Nemenyi post-hoc test~\cite{pohlert2014pairwise}. The outcome is represented by means of critical difference (CD) plots.

In each CD plot, the 15 competitors are placed on the horizontal axis according to their average ranking. Thereafter, the CD score is computed for a significance level of 0.05. If the difference is small enough, meaning that no statistical difference is detected, a horizontal bar linking statistically equal approaches is drawn. This analysis is done for all three levels of noise, and the corresponding CD plots are shown in Figures~\ref{fig:CDplots-no-noise}-\ref{fig:CDplots-noise0.01}.

@@

In this section we study the scalability of \textsc{RILS-ROLS} w.r.t. size of instances problems under different levels of noise.
Since we do not compare \textsc{RILS-ROLS} to other competitors here, but just assess the method scalability and sensitivity to noise, the experimental setup is different. First, we run each instance only once -- note that there are 5 randomly generated instances for each of 47 combinations of (size, variable count). Second, the exit criteria are different: 1) maximal running time is 300 seconds instead of 1 hour, and 2) maximal number of fitness evaluations is 100 thousands instead of 1 million. 
As before, the test set takes 25\% of total data. 
For the sake of having simpler presentation, we split the instances from \textsc{Random} benchmark into three parts as follows. 
\begin{itemize}
	\item \textit{Small-sized instances}: instances with solution size from 3 to 7.
	\item \textit{Medium-sized instances}: those with size from 8 to 11.
	\item \textit{Large-sized instances}: instances with size from 12 to 15. 
\end{itemize}

@@

	\item We can conclude that increased formula size and noise level jointly contribute to deterioration of overall \textsc{RILS-ROLS} accuracy, which is expected and natural. The diagram suggests that this deterioration behaves similar across different levels of noise when instance size changes from small to medium -- almost linearly, and with almost exact slope. On the other hand, the relative deterioration of accuracy under no-noise level, when going from medium to large instance is higher than under noisy data. Possible explanation for this is that the exact solution accuracy of medium instances is already very deteriorated in noisy scenario, therefore, further increase in size does not have significant impact. 
\end{itemize}

@@

The following conclusions may be drawn from these results. \fxnote{Marko: dosao dovde!}
